---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#|",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pegboard

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/pegboard)](https://CRAN.R-project.org/package=pegboard)
[![Travis build status](https://travis-ci.com/carpentries/pegboard.svg?branch=main)](https://travis-ci.com/carpentries/pegboard)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/carpentries/pegboard?branch=main&svg=true)](https://ci.appveyor.com/project/carpentries/pegboard)
[![Codecov test coverage](https://codecov.io/gh/carpentries/pegboard/branch/main/graph/badge.svg)](https://codecov.io/gh/carpentries/pegboard?branch=main)
[![R build status](https://github.com/carpentries/pegboard/workflows/R-CMD-check/badge.svg)](https://github.com/carpentries/pegboard/actions)
<!-- badges: end -->

> [pegboard] is tempered hardboard which is pre-drilled with evenly spaced
> holes. The holes are used to accept pegs or hooks to support various items,
> such as tools in a workshop.
>
> https://en.wikipedia.org/wiki/Pegboard

The {pegboard} package is a way to explore the Carpentries' lessons via their
XML representation. This package makes heavy use of rOpenSci's [{tinkr}](https://docs.ropensci.org/tinkr/) and [{xml2}](https://cran.r-project.org/package=xml2). 

There are two [{R6}](https://cran.r-project.org/package=R6) objects in the
package:

 - Episode: stores the xml content of a single episode
 - Lesson: stores all Episodes within a lesson

## Installation

This package is currently in development, but you can install it via {remotes}:

``` r
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
remotes::install_github("carpentries/pegboard")
```

## Example
 
The first way to get started is to use the `get_lesson()` function, which will
use [{git2r}](https://cran.r-project.org/package=git2r) to clone a lesson 
repository to your computer. 

```{r example, message = FALSE}
library(pegboard)
library(purrr)
library(xml2)
library(fs)

d <- fs::file_temp(pattern = "PBREADME")
rng <- get_lesson("swcarpentry/r-novice-gapminder", path = d)
rng

# Find all challenges
head(rng$challenges())

# Find all solutions
head(rng$solutions())

# Find all discussion blocks
head(rng$blocks(".discussion"))
```

## Manipulation

At the moment, you can manipulate each episode in various ways. One of the ways
that will become useful in the future is translating the episode from using the
old and busted Jekyll syntax (e.g. using nested block quotes to create
specialized sections and writing questions/keypoints/objectives in the YAML) to 
using a more intuitive system (currently being evaluated). For example, let's
say we wanted to transform an episode from Jekyll to using
[{sandpaper}](https://github.com/zkamvar/sandpaper#readme). This would involve
the following steps:

1. transforming the block quotes to native or fenced div tags (or dovetail blocks)
1. converting code block decorators (`{: .language-r}`) and modify setup chunk
1. moving questions, objectives, and keypoints to the body of the document

Doing this by hand would be a nightmare, but we've written {pegboard} in such a
way that will streamline this process

First, let's inspect how the file looks at the moment:

```{r inspect the file}
fun <- rng$episodes$`10-functions.Rmd`
fun$write(d, format = "Rmd")
cat(readLines(fs::path(d, fun$name), n = 60), sep = "\n")
```

Now, we can apply the transformation chain in the order we specifed:

```{r chain}
fun$
  unblock()$         # transform block quotes
  use_sandpaper()$   # convert code block decorators and modify setup chunk
  move_questions()$  # ...
  move_objectives()$
  move_keypoints()$
  write(d, format = "Rmd")
cat(readLines(fs::path(d, fun$name), n = 70), sep = "\n")
```

## Reset

All changes can be reset to the initial state with the `$reset()` method:

```{r reset}
fun$reset()
cat(readLines(fs::path(d, fun$name), n = 70), sep = "\n")
```

