---
title: "Introduction to the Lesson Object"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the Lesson Object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##"
)
```

## Introduction

```{r child="../man/rmd-fragments/intro-class.Rmd"}
```

If you have a list of `Episode` objects, you can achieve _most_ of what you can
with the lesson objects, because much of the `Lesson` object's function is to
provide a methods that map over all of the `Episode` object methods. The key
difference between `Lesson` objects and a list of `Episode` objects is that
the `Lesson` object will collapse summaries and to map relations between 
`Episodes` and their children or parent documents.

**Before you read this vignette, please read the vignette on the `Episode`
object (`vignette("intro-episode", package = "pegboard")`)** so that you can
understand the methods that come from the `Episode` objects. In this vignette,
we will talk about the structure of `Lesson` objects, how to use the basic
methods of these objects, inspecting summaries of source vs built episodes, and
assessing the lineage of episodes that have parents and/or children documents.

But first, because of a default parameter that influences what methods can be
used depending on the lesson context, I need to explain a little bit about
Jekyll, the former lesson infrastructure.

### A Brief Word About History and Jekyll

Prior to The Workbench, we had the [styles
repository](https://github.com/carpentries/styles/), which was an all-in-one
toolbox that built websites with Jekyll. It was colloquially known as the 
"Lesson Template." It has two major differences to The Workbench: folder 
structure and syntax.

#### Folder Structure

The folder structure of lessons built with Jekyll was one where content and
tooling lived side-by-side. This folder structure looked something like this:


```{r, jekyll-folders, echo = FALSE, comment = "# "}
writeLines(".
├── Gemfile
├── Makefile
├── _config.yml
├── \033[01;34m_episodes/\033[0m
│   └── 01-intro.md
├── \033[01;34m_episodes_rmd/\033[0m
├── \033[01;34m_extras/\033[0m
│   ├── a.md
│   └── b.md
├── \033[01;34m_includes/\033[0m
├── \033[01;34m_layouts/\033[0m
├── aio.md
├── \033[01;34massets/\033[0m
├── \033[01;34mbin/\033[0m
├── \033[01;34mfig/\033[0m
├── index.md
├── reference.md
├── requirements.txt
└── setup.md
")
```

When {pegboard} was first written, we initially assumed this folder structure,
where R Markdown and regular Markdown episodes lived in different folders (and
more often than not, the outputs of the R Markdown files lived inside the 
`_episodes/` folder. The main method of organising episodes was by numbers
embedded in the name of the files.

As The Workbench developed, it was clear that this folder structure needed to
change, but we needed to keep compatibility with the old lessons because we
want to ensure that people can independently convert from the old style lessons
to the new style, thus we added the `jekyll` parameter to the `Lesson` object
initializer method, and set `jekyll = TRUE` as the default to keep backwards
compatibility.


## Creating a New Lesson Object

The `Lesson` object is invoked with the `Lesson$new()`, method. Here, I will
demonstrate a Workbench lesson. This is the folder structure of the workbench
lesson:

```{r intro-read-noshow, echo = FALSE, comment = "# "}
withr::with_dir(pegboard::lesson_fragment("sandpaper-fragment"), {
  fs::dir_tree(regex = "site/[^R].*", invert = TRUE)
})
```

To read it in, because we have a Workbench lesson, we need to specify `jekyll =
FALSE` to register all the div tags and ensure that the lesson is being treated
like a Workbench lesson.

```{r setup}
library("pegboard")
library("glue")
library("yaml")
library("xml2")
library("fs")
wbfragment <- lesson_fragment("sandpaper-fragment")
print(wbfragment) # path to the lesson
wb_lesson <- Lesson$new(wbfragment, jekyll = FALSE)
print(wb_lesson)
```

The Lesson printing here shows that it has a subset of methods that are named
similarly to methods and active bindings from the Episode class. These are not
inherited, but rather they are implemented across all the Episode objects. The
Episode objects themselves are parsed into one of three elements: "episodes",
"children", and "extra" (NOTE: the "extra" slot may be superceded by elements 
that better match the folder structure of lessons).

```{r show-episodes}
lapply(wb_lesson$episodes, class)
lapply(wb_lesson$children, class)
lapply(wb_lesson$extra, class)
```

Notice here that there is only one episode in the `$episodes` item, but in the
directory tree above, we see two. This is because of the `config.yaml` file,
which defines the order of the episodes:

```{r config-order}
read_yaml(path(wbfragment, "config.yaml"))$episodes
```

Because `episodes/nope.Rmd` is not listed, it is not read in. This is useful
to avoid reading in content from files that are incomplete or not correctly
formatted. 

## File Information

The Lesson object contains information about the file information:

```{r show-files}
# what is the root path for the lesson?
wb_lesson$path
# what episode files exist?
wb_lesson$files
# do any of the files have children (Workbench lessons only)? 
wb_lesson$has_children
```

## Accessors

As mentioned earlier, many of the methods in a `Lesson` object are wrappers
for methods in `Episode` objects. `challenges`, `solutions` are the obvious
ones. 

```{r challenge-solution}
wb_lesson$challenges()
wb_lesson$solutions()
```

For the rest of the elements (or active bindings), you will need to use the
`$get()` method. For example, if you wanted all code blocks from the episodes
and the extra content, you would use:

```{r get-code}
wb_lesson$get("code", collection = c("episodes", "extra"))
```

Similarly, for links and headings you would use: 

```{r get-links-headings}
wb_lesson$get("links", collection = c("episodes", "extra"))
wb_lesson$get("headings", collection = c("episodes", "extra"))
```

## Methods Summaries and Validation

For summaries, you will get a data frame of the summaries. You can also choose
to include other collections in the summary:

```{r summary}
wb_lesson$summary() # defaults to episodes
wb_lesson$summary(collection = c("episodes", "extra"))
```

Validation will auto-check everything and return the results as data frames. You can find more information abou the specific checks by reading `vignette("validation", package = "pegboard")`. 

Details of the individual functions can be found via `?validate_links()`,
`?validate_divs()`, and `?validate_headings()`.

```{r validate}
divs <- wb_lesson$validate_divs()
print(divs)
headings <- wb_lesson$validate_headings()
print(headings)
links <- wb_lesson$validate_links()
print(links)
```

## Loading Built Documents

One thing that is very useful is to check the status of the built documents
to ensure that everything you expect is there. You can load all of the built markdown documents with the `$load_built()` method and the built documents will
populate the `$built` field:

```{r load-built}
wb_lesson$load_built()
lapply(wb_lesson$built, class)
```

You can use these to inspect how the content is rendered and see that the
code blocks render what they should render. In thise case, `episodes/intro.Rmd`
will render one output block and one image.

```{r summary-built}
to_check <- c("page", "code", "output", "images", "warning", "error")
wb_lesson$summary(collection = c("episodes", "built"))[to_check]
```

## Handouts

This is another method wrapped from the Episode method, where it combines the
output into a single file and prepends the Episode title before each section:

```{r handouts}
writeLines(wb_lesson$handout())
```

## Accessing other `Episode` methods

For `pegboard::Episode` methods that are not listed above, you will need to
manually iterate over the `Episode` objects. For example, if you wanted to
extract all of the instructor notes in the lesson, you could use `purrr::map()`

```{r callouts}
purrr::map(c(wb_lesson$episodes, wb_lesson$extra), 
  function(ep) ep$get_divs("instructor"))
```

If you wanted to get a specific thing from the body of the document, then you
could use any of the functions from {xml2} such as `xml2::xml_find_first()` or
`xml2::xml_find_all()`. Here, we are looking first the first text element that
is not a fenced-div element:

```{r body-text}
purrr::map_chr(c(wb_lesson$episodes, wb_lesson$extra), 
  function(ep) {
    xpath <- ".//md:text[not(starts-with(text(), ':::'))]"
    nodes <- xml_find_first(ep$body, xpath, ep$ns)
    return(xml_text(nodes))
  }
)
```

For more information about constructing XPath queries and working with XML data,
you can read `vignette("intro-xml", package = "pegboard")`


## Creating a New Lesson with Child Documents

If you are unfamiliar with the concept of child documents, please read
the "Including Child Documents" vignette in the {sandpaper} package
(`vignette("include-child-documents", package = "sandpaper")`).

The `pegboard::Lesson` object is very useful with lessons that contain child
documents because it records the relationships between documents. This is key
for workflows determining build order of a Lesson. If a source document is
modified, in any build system, that source document will trigger a rebuild of
the downstream page, and _the same should happen if a child document of that
source is modified_ (if you are interested in the build process used by
{sandpaper}, you can read `sandpaper::build_status()` and 
`sandpaper::hash_children()`). This functionality is implemented in the
`pegboard::Lesson$trace_lineage()` method, which returns all documents required
to build any given file. We will demonstrate the utility of this later, but
first, we will demonstrate how `pegboard::Lesson$new()` auto-detects the child
documents:

Take for example the same lesson, but `episodes/intro.Rmd` has the child
`episodes/files/cat.Rmd` which in turn has the child
`episodes/files/session.Rmd`:

```{r intro-read-noshow-children, echo = FALSE, comment = "# "}
withr::with_dir(lesson_fragment("sandpaper-fragment-with-child"), {
  fs::dir_tree(regex = "site/[^R].*", invert = TRUE)
})
```

A valid child document reference requires a code chunk with a `child` attribute
that points to a valid file relative to the parent document, so if I have this
code block inside `episodes/intro.Rmd`, then it will include the child document
called `episodes/files/cat.Rmd`:

````{verbatim}

```{r cat-child, child="files/cat.Rmd"}
```
````

During initialisation of a Workbench lesson (note: not currently implemented for
Jekyll lessons), the `Lesson` object will detect that at least one `Episode`
references at least one child document (via `find_children()`) and read them in
(see `load_children()` for details).

```{r show-children-files}
wbchild <- lesson_fragment("sandpaper-fragment-with-child")
wb_lesson_child <- Lesson$new(wbchild, jekyll = FALSE)
wb_lesson_child$has_children
lapply(wb_lesson_child$children, class)
```

The reason it is useful is because if you have a child Episode object, you can
determine its parent and its final ancestor. Because these paths are absolute
paths, I am going to write a function that will use the {glue} package to print
it nicely for us: 

```{r show-parent}
show_child_parents <- function(child) {
  parents <- fs::path_rel(child$parents, start = child$lesson)
  build_parents <- fs::path_rel(child$build_parents, start = child$lesson)

  msg <- "Ancestors for {child$name} ---
  Parent(s):         {parents}
  Final ancestor(s): {build_parents}"
  glue::glue(msg)
}

# cat.Rmd's parent is intro.Rmd
show_child_parents(wb_lesson_child$children[[1]])

# session.Rmd's parent is cat.Rmd
show_child_parents(wb_lesson_child$children[[2]])
```

If you have the name of the final ancestor, then you can determine the full
lineage with the `$trace_lineage()` method, which is useful for determining if
a file should be rebuilt:

```{r show-lineage}
parent <- wb_lesson_child$children[[2]]$build_parents
print(parent)
lineage <- wb_lesson_child$trace_lineage(parent)

# printing the lineage in a presentable fashion:
rel <- wb_lesson_child$path
pretty_lineage <- path_rel(lineage, start = rel)
pretty_lineage <- glue_collapse(pretty_lineage, sep = ", ", last = ", and ")
glue("The lineage of {path_rel(parent, start = rel)} is:
  {pretty_lineage}")
```


## Jekyll Lessons

This section will talk about the peculiarities with lessons built with the
[carpentries/styles](https://carpentries.github.io/lesson-example/) lesson
template. Note that lesson transition methods are not implemented in the
`Lesson` object, if you want to find out about methods for transition, please
read the Jekyll section in `vignette("intro-episode", package = "pegboard")`.


### Syntax

```{r child="../man/rmd-fragments/jekyll-syntax.Rmd"}
```

### Methods

There are some methods that are specific to lessons that are built with Jekyll.
In particular, the `n_problems` and `show_problems` active binding are useful
for determining if anything went wrong with parsing the kramdown syntax,
the `$isolate_blocks()` method was used to strip out all non-blockquote content,
the `$blocks()` method returned all block quote with filters for types, and
the `$rmd` field was an indicator that the lesson used R Markdown.


```{r jekyll}
jekyll <- Lesson$new(lesson_fragment("lesson-fragment"), jekyll = TRUE)
jekyll$n_problems
rmd    <- Lesson$new(lesson_fragment("rmd-lesson"), jekyll = TRUE)
rmd$n_problems
```

As mentioned above, in Jekyll uses special block quotes to format callout 
blocks. The `$challenges()` and `$solutions()` methods recognise this and will
return the block quotes:

```{r jeky-chal}
jekyll$challenges()
jekyll$solutions()
```

For other blocks, you can use the `$blocks()` method:

```{r jeky-blocks}
jekyll$blocks(".prereq")
rmd$blocks(".prereq")
```






