---
title: "Introduction to the Episode Object"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the Episode Object}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The {pegboard} package facilitates the analysis and manipulation of Markdown and
R Markdown files by translating them to XML and back again. This extends the
{tinkr} package by providing additional methods that are specific for 
Carpentries-style lessons. There are two `R6` classes defined in {pegboard}:

 - `Episode` objects that contain the XML data, YAML metadata and extra fields
   that define the child and parent files for a particular episode
 - `Lesson` objects that contain lists of `Episode` objects categorised as 
   "episodes", "extra", or "children"

This vignette will be discussing the structure of Episode objects, how to 
query the contents with the {xml2} package, and how to use the methods and 
active bindings to get information about, extract, and manipulate anything
inside of a Markdown or R Markdown document.

## Reading Markdown Content

Each `Episode` object starts from a Markdown file. In particular for {pegboard},
we assume that this Markdown file is written using
[Pandoc](https://pandoc.org/MANUAL.html) syntax (a superset of
[CommonMark](https://commonmark.org/)). It can be any markdown file, but for us
to explore what the `Episode` object has to offer us, let's take an example R
Markdown file that is present in a fragment of a Carpentries Workbench lesson
that we have in this package. We will be using the {xml2} package to explore
the object and the {fs} package to help with constructing file paths.

```{r setup}
library("pegboard")
library("xml2")
library("fs")
```

This is what our lesson fragment looks like. It is a fragment because it's main
purpose is to be used for examples and tests, but it contains the basic structure
of a lesson that we want. 

```{r intro-read-noshow, echo = FALSE}
dir_tree(lesson_fragment("sandpaper-fragment"), recurse = 1, regex = "site/[^R].*", invert = TRUE)
```

We can retrieve it with the `lesson_fragment()` function, which loads example
data from pegboard. Here we will take that lesson fragment and read in the first
episode with the initialization method, `Episode$new()`, followed by
`$confirm_sandpaper()`, a confirmation that the episode was created to work
with [{sandpaper}], the user interface and build engine of The Carpentries
Workbench (for information on non-workbench content, see the section on [Jekyll
Lesson Markdown Content](#jekyll-lesson-markdown-content)) and `$protect_math()`
which will prevent special characters in LaTeX math from being escaped.

[{sandpaper}]: https://carpentries.github.io/sandpaper/

```{r intro-read}
lsn <- lesson_fragment("sandpaper-fragment")
# Read in the intro.Rmd document as an `Episode` object
intro_path <- path(lsn, "episodes", "intro.Rmd")
intro <- Episode$new(intro_path)$confirm_sandpaper()$protect_math()
```

If we print out the Episode object, I'm going to get a long list of methods,
fields and active bindings (functions that act like fields) printed:

```{r intro-print}
intro
```

The actual XML content is in the `$body` field. This contains all the data from
the markdown document, but in XML form. 

```{r intro-body}
intro$body
```

If we want to see what the contents look like, you can use the `$show()`,
`$head()`, or `$tail()` methods (note: the `$show()` method will print out the
entire markdown document).

```{r intro-show}
intro$head(10)
intro$tail(10)
intro$show()
```

## File information

For information about the file and its relationship to other files, you can use
the following active bindings, which are useful when working with Episodes in a
lesson context.

```{r file-active-bindings}
intro$path
intro$name
intro$lesson
# NOTE: relationships to other episodes are automatically handled in the
#       Lesson context
intro$has_parents
intro$has_children
intro$children # separate documents processed as if they were part of this document
intro$parents  # the immediate documents that would require this document to build
intro$build_parents # the final documents that would require this document to build
```

## Accessing Markdown Elements

The `Episode` object is centered around the `$body` item, which contains the XML
representation of document. It is possible to find markdown elements from XPath
statments:

```{r xpath-active-bindings}
xml2::xml_find_all(intro$body, ".//md:link", ns = intro$ns)
xml2::xml_find_first(intro$body, ".//md:list[@type='ordered']", ns = intro$ns)
```

However, there are some useful elements that we want to know about, so I have
implemented them in active bindings and methods:


```{r active-bindings}
# headings where level 2 headings are equivalent to sections
intro$headings
# all callouts/fenced divs
intro$get_divs()
intro$challenges
intro$solutions
# questions, objectives, and keypoints are standard and return char vectors
intro$objectives 
intro$questions
intro$keypoints
# code blocks and output types
intro$code
intro$output
intro$warning
intro$error
# images and links
intro$images
intro$get_images() # parses images embedded in `<img>` tags
intro$links
```

Much of these are summarized in the `$summary()` method:

```{r summary}
intro$summary()
```

## Code blocks and code chunks

In markdown, a **code block** is written with fences of at least three backtick
characters (`` ` ``) followed by the language for syntax highlighting:

````markdown

List all files in reverse temporal order, printing their sizes in
a human-readable format:

```bash
ls -larth /path/to/folder
```
````

> List all files in reverse temporal order, printing their sizes in
> a human-readable format:
> 
> ````bash
> ls -larth /path/to/folder
> ````

When these are processed by {pegboard}, the resulting XML has this structure
where the backticks inform that kind of node (`code_block`) and the language
type is known as the "info" attribute. Everything inside the code block is the
node text and has whitespace preserved

````{r show-code-block, echo = FALSE, results = 'asis'}
cb <- "```bash

ls -larth /path/to/folder
```"
cbx <- xml2::read_xml(commonmark::markdown_xml(cb))
txt <- as.character(xml2::xml_find_first(cbx, ".//d1:code_block"))
writeLines(c("```xml", txt, "```"))
````

In R Markdown, there are special code blocks that are called code chunks that
can be dynamically evaluated. These are distinguished by the curly braces
around the language specifier and [optional
attributes](https://yihui.org/knitr/options/) that control the output of the
chunk. 

````{verbatim}

There is a code chunk here that will produce a plot, but not show the code:

```{r chunky, echo=FALSE, fig.alt="a plot of y = mx + b for m = 1 and b = 0"}
plot(1:10, type = "l")
```

````


> There is a code chunk here that will produce a plot, but not show the code:
>
> ````{r chunk-name, echo = FALSE, fig.alt="a plot of y = mx + b for m = 1 and b = 0"}
> plot(1:10, type = "l")
> ````

When this is processed with {pegboard}, the "info" part of the code block is
further split into "language", "name" and further attributes based on the chunk
options:

````{r show-code-chunk, echo = FALSE, results = 'asis'}

chunk <- 'There is a code chunk here that will produce a plot, but not show the code:

```{r chunky, echo=FALSE, fig.alt="a plot of y = mx + b for m = 1 and b = 0"}

plot(1:10, type = "l")
```'
tmp <- tempfile()
writeLines(chunk, tmp)
chunky <- pegboard::Episode$new(tmp)$code[[1]]
xml2::xml_set_attr(chunky, "sourcepos", NULL)
txt <- as.character(chunky)
writeLines(c("```xml", txt, "```"))
unlink(tmp)
````

Both code blocks will be encountered, but the difference between them is that
the R Markdown code chunks will have the "language" attribute. This is an
important concept to know about when you are searching and manipulating R 
Markdown documents with XPath 
(see `vignette("intro-xml", package = "pegboard")`). The next section will walk
through some aspects of manipulation that we can do with these documents.

## Manipulation

Because everything centers around the `$body` element and is extracted with 
{xml2}, it's possible to manipulate the elements of the document. One thing that
is possible is that we can add new content to the document using the `$add_md()`
method, which will add a markdown element after any paragraph in the document.

For example, we can add information about pegboard with a new code block after 
the first heading:

````{r add-code-block}
intro$head(26) # first 26 lines
intro$body # first heading is item 11
cb <- c("You can clone the **{pegboard} package**:

```sh
git clone https://github.com/carpentries/pegboard.git
```
")
intro$add_md(cb, where = 11)
intro$head(26) # code block has been added
intro$code
````

You can also manipulate existing elements. For example, let's say we wanted to
make sure all R code chunks were named. We can do so by querying and
manipulating the code blocks:

```{r update-code-block}
code <- intro$code
code
# executable code chunks will have the "language" attribute
is_chunk <- xml2::xml_has_attr(code, "language")
chunks <- code[is_chunk]
chunk_names <- xml2::xml_attr(chunks, "name")
nonames <- chunk_names == ""
chunk_names[nonames] <- paste0("chunk-", seq(sum(nonames)))
xml2::xml_set_attr(chunks, "name", chunk_names)
code
```

We can see that the chunks now have names, but the proof is in the rendering:

```{r show-updated}
intro$show()
```

One of the things about manipulating these documents in code is that it is 
possible to go back and reset if things are not correct, which is why we have
the `$reset()` method:

```{r}
intro$reset()$confirm_sandpaper()$protect_math()$head(25)
```

## Jekyll Lesson Markdown Content

This section describes the features that you would expect to find in a lesson
that was built with the former infrastructure,
<https://github.com/carpentries/styles>, which was built using the Jekyll
static site generator. These style lessons are no longer supported by The
Carpentries. {pegboard} does support these lessons so that they can be 
transitioned to use The Workbench syntax via [The Carpentries Lesson Transition
Tool](https://github.com/carpentries/lesson-transition#readme).

### The Syntax of Jekyll Lessons

The former Jekyll syntax used [kramdown-flavoured
markdown](https://kramdown.gettalong.org/syntax.html), which evolved separately
from [commonmark](https://spec.commonmark.org/), the syntax that {pegboard} 
knows and that Pandoc-flavoured markdown extends. One of the key differences 
with the kramdown syntax is that it used something known as [Inline Attribute 
Lists (IAL)](https://kramdown.gettalong.org/syntax.html#inline-attribute-lists) to
help define classes for markdown elements. These elements were formated as
`{: <attributes>}` where `<attributes>` is replaced by class definitions and
key/value pairs. They always appear _after_ the relevant block which lead to
code blocks that looked like this:

````markdown
~~~
ls -larth /path/to/dir
~~~
{: .language-bash}
````

Moreover, to achieve the special callout blocks, we used blockquotes that were
given special classes (which is an accessbility no-no because those blocks were
not semantic HTML) and the nesting of these block quotes looked like this:


````markdown
> ## Challenge
> 
> How do you list all files in a directory in reverse order by the time it was 
> last updated?
> 
> > ## Solution
> > 
> > ~~~
> > ls -larth /path/to/dir
> > ~~~
> > {: .language-bash}
> {: .solution}
{: .challenge}
````

One of the biggest chalenges with this for authors was that, unless you used an
editor like vim or emacs, this was difficult to write with all the prefixed 
blockquote characters and keeping track of which IALs belonged to which block.


