---
title: "Software Carpentry Survey"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

I wrote this package to help me parse the Carpentries' markdown files so that
I can understand how I want to update the lesson templates. 

```{r setup}
library(tidygraph)
library(magrittr)
library(pegboard)
library(ggraph)
library(dplyr)
library(purrr)
library(glue)
library(xml2)
library(gt)
```


```{r download_swc_lessons, results = "hide"}
swc_path <- fs::file_temp(pattern = "SWC")
swc_core_lessons <- c(
  "shell-novice",
  "shell-novice-es",
  "git-novice",
  "git-novice-es",
  "python-novice-inflammation",
  "python-novice-gapminder",
  "r-novice-inflammation",
  "r-novice-gapminder",
  "r-novice-gapminder-es",
  NULL
)
SWC <- map(glue("swcarpentry/{swc_core_lessons}"), get_lesson, path = swc_path)
names(SWC) <- swc_core_lessons
```

One of the things I like to look at is what unique patterns we find with the
block elements in the challenges. We can use the `graph = TRUE` element to check
this across our curricula. Setting `recurse = TRUE` dips into the block elements
that are nested within block elements.

```{r curr-graph}
relationships <- map_dfr(SWC, ~.x$challenges(graph = TRUE, recurse = TRUE), .id = "Lesson") %>% as_tibble()

relationships

curr_graph <- relationships %>%
  dplyr::mutate(level = dplyr::case_when(
    level == 1 ~ "Challenge", 
    level == 2 ~ "Solution"
  )) %>%
  dplyr::mutate(grp = Lesson) %>%
  dplyr::group_by(grp) %>% 
  dplyr::distinct(Lesson, Episode, from, to, level) %>%
  dplyr::group_map(~as_tbl_graph(.x))

curr_graph[[1]]
```

```{r, show-graph, fig.width = 12, fig.height = 12}
walk(
  curr_graph, 
  ~print(
    ggraph(.x, layout = 'kk') + 
      geom_edge_fan(
        aes(alpha = stat(index), color = level), 
        arrow = grid::arrow(length = ggplot2::unit(0.125, "inches"))
      ) + 
      geom_node_label(aes(label = name)) +
      facet_edges(~Episode) + 
      theme_graph(foreground = 'steelblue', fg_text_colour = 'white') +
      ggtitle(as_tibble(activate(.x, edges))$Lesson[1]) +
      NULL
  )
)
```


I wasnt to examine the relationships in a bit finer detail. The way the graph
works is that the solution nodes will either point to the contents inside the
solution block (normally the header), or the exit point from the solution block
(either an element from the challenge block or out to the lesson itself).

How many ways do we enter and exit solutions?

```{r solutions, warn = FALSE}
n_blocks <-  relationships %>%
  group_by(Lesson, Episode, Block) %>%
  count() %>%
  nrow()

res <- relationships %>%
  filter(from == "solution") %>%
  count(from, to, level) %>%
  arrange(level, desc(n)) %>%
  mutate(fraction = n/n_blocks)

res %>%
  select(level, exit = to, n, percent = fraction) %>%
  mutate(exit = gsub("_", " ", exit)) %>%
  mutate(exit = `substring<-`(exit, 1, 1, toupper(substr(exit, 1, 1)))) %>%
  gt() %>%
  tab_row_group(
    "...enters at",
    rows = level == 2
  ) %>%
  tab_row_group(
    "...exits to",
    rows = level == 1
  ) %>%
  fmt_percent(
    columns = vars(percent),
    decimals = 2
  ) %>%
  fmt_number(
    columns = vars(n),
    decimals = 0
  ) %>%
  cols_hide(columns = vars(level)) %>%
  cols_align(
    columns = vars(exit),
    align = "left"
  ) %>%
  cols_align(
    columns = vars(n),
    align = "right"
  ) %>%
  cols_width(
    vars(exit) ~ px(140),
    everything() ~ px(80)
  ) %>%
  cols_label(
    exit    = md("A **solution block**"),
    n       = md("**n**"),
    percent = md("**%**")
  ) %>% 
  opt_row_striping() 
```

What I can tell from here is that % of the time, solutions will dump out into
the lessons, but other times the challenge block may continue. There is the 
weird bit where there are solutions that go to solutions. What's going on there?

```{r to_solutions}
relationships %>%
  filter(from == "solution", to == "solution") %>%
  select(Lesson, Episode, pos, everything())

# Adding ten lines to the range and the length of the yaml
readLines(SWC$`python-novice-inflammation`$files[2])[(615:(638 + 10)) + 23]
readLines(SWC$`python-novice-inflammation`$files[3])[(129:(137 + 10)) + 13]
readLines(SWC$`python-novice-inflammation`$files[7])[(291:(302 + 10)) + 20]
```

```{r non-lesson-exits}
relationships %>%
  filter(
    from == "solution",
    to != "solution",
    to != "lesson",
    level == 1
  ) %>% 
  select(-level, -from) %>%
  mutate(Block = as.integer(Block)) %>% 
  distinct(Lesson, Episode, Block) -> odd_blocks
odd_blocks 
odd_blocks %<>%
  dplyr::rowwise() %>%
  mutate(src = list(SWC[[Lesson]]$challenges()[[Episode]][[Block]]))

names(odd_blocks$src) <- glue_data(odd_blocks, 
  "{Lesson}/{Episode} (Block {Block})"
)
purrr::map(SWC, ~.x$episodes) %>% 
  purrr::flatten() %>%
  purrr::walk(~.x$use_dovetail()$unblock())
purrr::walk2(names(odd_blocks$src), odd_blocks$src, ~cat(.x, xml_text(.y), sep = "\n---\n"))
```

