---
title: "swc-survey"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{swc-survey}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

I wrote this package to help me parse the Carpentries' markdown files so that
I can understand how I want to update the lesson templates. 

```{r setup}
library(tidygraph)
library(magrittr)
library(up2code)
library(ggraph)
library(purrr)
library(glue)
library(xml2)
```


```{r download_swc_lessons, results = "hide"}
swc_path <- fs::file_temp(pattern = "SWC")
swc_core_lessons <- c(
  "shell-novice",
  "shell-novice-es",
  "git-novice",
  "git-novice-es",
  "python-novice-inflammation",
  "python-novice-gapminder",
  "r-novice-inflammation",
  "r-novice-gapminder",
  "r-novice-gapminder-es",
  NULL
)
SWC <- map(glue("swcarpentry/{swc_core_lessons}"), get_lesson, path = swc_path)
names(SWC) <- swc_core_lessons
```

```{r swc_explore}
map_int(SWC, ~length(.x$episodes))
map(SWC, ~.x$challenges() %>% lengths())
```

One of the things I like to look at is what unique patterns we find with the
block elements in the challenges. We can use the `graph = TRUE` element to check
this across our curricula. Setting `recurse = TRUE` dips into the block elements
that are nested within block elements.

```{r curr-graph}
curr_graph <- map_dfr(SWC, ~.x$challenges(graph = TRUE, recurse = TRUE), .id = "Lesson") %>%
  dplyr::mutate(level = dplyr::case_when(
    level == 1 ~ "Challenge", 
    level == 2 ~ "Solution"
  )) %>%
  dplyr::mutate(grp = Lesson) %>%
  dplyr::group_by(grp) %>% 
  dplyr::distinct(Lesson, Episode, from, to, level) %>%
  dplyr::group_map(~as_tbl_graph(.x))

curr_graph
```

```{r, show-graph, fig.width = 12, fig.height = 12}
walk(
  curr_graph, 
  ~print(
    ggraph(.x, layout = 'kk') + 
      geom_edge_fan(
        aes(alpha = stat(index), color = level), 
        arrow = grid::arrow(length = ggplot2::unit(0.125, "inches"))
      ) + 
      geom_node_label(aes(label = name)) +
      facet_edges(~Episode) + 
      theme_graph(foreground = 'steelblue', fg_text_colour = 'white') +
      ggtitle(as_tibble(activate(.x, edges))$Lesson[1]) +
      NULL
  )
)
```

