---
title: "Validation of Lesson Elements"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Validation of Lesson Elements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#|",
  R.options = list(width = 100, str = strOptions(strict.width = "cut"))
)
```

## Introduction

This package is designed to parse the structure of the Markdown and R Markdown
files inside a Carpentries-style lesson. To ensure consistent lesson structure,
it has built-in validators that will inspect headings, links, and callout
sections. This document will help you become familiar with the output of the
validators and how they can be used to update a lesson.

Lesson authors and contributors will likely only see the output of these
validators and not need to interact with them directly. For example, we have a
lesson that contains some mal-formed links and headings:

```{r setup}
library(pegboard)
lsn <- Lesson$new(lesson_fragment())
# validation
lsn$validate_links()
lsn$validate_headings(verbose = FALSE)
```

We can see that there is informative and actionable output produced that will
allow lesson authors to make targeted changes in their files. The next sections
will detail how we can work with the output of these methods.

## Link Validation

Each of the validation methods will produce messages (aka stderr) for the user,
but they also produce a data frame as output that contains detailed information
about each link and what tests it passed or did not pass. You can find out what
tests are run on links by accessing the help page for `validate_links` by typing 
`?validate_links` in your R console.

If we want to inspect this data frame, we can assign it to a new variable:

```{r validation1}
links <- lsn$validate_links()
str(links, max.level = 1)
```

This data frame is combined output of three sources:

1. the output of [xml2::url_parse()]
2. the source data containing the original source file (filepath), line number
   (pos), and XML node (node)
3. the tests performed as logical vectors, which can be extracted programmatically

Here, we can see the nodes that did not pass our tests by using the `link_tests`
vector, which is an internal vector defining the inline messages printed for 
each failed test, to filter the output:

```{r show-invalid-links}
# get the subset of rows that did not pass all the tests
invalid <- !apply(links[names(link_tests)], MARGIN = 1L, all)
# return the nodes
links$node[invalid]
```

For context, this is what the document looks like at those positions:

```{r tail-link, echo = FALSE}
lsn$episodes[["14-looping-data-sets.md"]]$tail(17)
```

### Headings

Validation of headings operate similarly to links as it produces a data frame
along with the message output that can be further inspected and manipulated. You
can find out more by accessing the help documentation for `validate_headings` by
typing `?validate_headings` in your R console.

```{r heading-validation}
headings <- lsn$validate_headings(verbose = TRUE)
str(headings, max.level = 1)
```

This particular data frame has fewer rows because there are fewer moving parts
to headings than links, but they are indeed important. The process for getting 
the subset of invalid headings is similar: use the `heading_tests` vector from
pegboard to subset the rows that failed:


```{r show-invalid}
# get the subset of rows that did not pass all the tests
invalid <- !apply(headings[names(heading_tests)], MARGIN = 1L, all)
# return the nodes
headings$node[invalid]
```


## Internal testing

For testing, we have created some documents that have extreme cases of links and
headings that failed to show all the features of our examples:

```{r show-test-examples}
ex <- system.file("tests/testthat/examples", package = "pegboard")
lnk <- Episode$new(fs::path(ex, "link-test.md"))
hd  <- Episode$new(fs::path(ex, "validation-headings.md"))
```

### Links

The links example has _several_ mistakes covering our possibilities:

```{r show-link-test}
lnk$show()
lnk$validate_links()
```

### Headings

The headings example also has several mistakes, and demonstrates the value of
having a visual heading tree displayed on output when `verbose = TRUE`

```{r show-heading-test}
hd$show()
hd$validate_headings(verbose = TRUE)
```



