---
title: Basic Examples
output:
  rmarkdown::md_document:
    variant: gfm
---

```{r}
knitr::opts_chunk$set(error = TRUE)
```


These are some basic examples of syntax we might expect for learner.

On the face of it, one should be able to construct blocks with a single
rmarkdown chunk that has special tags embedded. This way, the RMarkdown is valid
no matter what.

## The default

This example will use the default settings from the document to render the
exercise.

```{r exercise-default}
#' @exercise Test Exercise
#' This is the text for an exercise.
#'
#' There might be some text and code blocks explaining things
#'
#' ```bash
#' Rscript -e 'plot(rnorm(100))'
#' ```
## there might be code as part of the exercise.
plot(rnorm(100))

#' @solution
#' the solution might include text like this line
#’
## or code

plot(rnorm(100), type = "b")
```

## Unworkable examples

Sometimes the example shown should not be worked. If we run this normally, R
will throw an error and it will look very messy. However, if we were to specify
`eval = FALSE` for the whole chunk, then the solution would not be run. To
alleviate this, we try to introduce tag-level evaluation by combining roxygen
syntax with chunk syntax.

```{r exercise}
#' @exercise {eval=FALSE}
#' This is the text for an exercise.

## there might be code as part of the exercise.
## but that is not working
plot(some_undefined_variable)

#' @solution {eval=TRUE}
#' the solution might include text like this that would explain
#’ why the code doesn’t work
## or code

exists("some_defined_variable")
```

## Wizard behind the curtain

On persistant form from R-novice gapminder is to include the solution in the
example, but only show the output or the last line.

```{r, wizard}
#' @exercise {echo=-1}
#' Using `seq`, write a function that prints the first **N** natural numbers, one per line:
#'
print_N <- function(N) {
 nseq <- seq(N)
 for (num in nseq) {
   print(num)
 }
}
print_N(3)
#'
#' @solution
#'
print_N <- function(N) {
 nseq <- seq(N)
 for (num in nseq) {
   print(num)
 }
}
```

## Multiple examples

However, there is a catch that happens whenever there are multiple examples per
block: some of them need to be echoed and others don't necessarily need it.

In these cases, we can have the user re-add the tag to specify new options that
would be persistent until the next tag.

```{r, multi-wizard}
#' @exercise
#' R has a built-in function called `seq` that creates a list of numbers:
#'
seq(3)
#'
#' Using `seq`, write a function that prints the first **N** natural numbers, one per line:
#' @exercise {echo=-1}
## This example only shows the very last line
print_N <- function(N) {
 nseq <- seq(N)
 for (num in nseq) {
   print(num)
 }
}
print_N(3)
#'
#' @solution
print_N <- function(N) {
 nseq <- seq(N)
 for (num in nseq) {
   print(num)
 }
}
```
