---
title: "Example Blocks from Carpentries Lessons"
authors: ["Zhian N. Kamvar", "François Michonneau"]
---

This document contains example/solution blocks from carpentries lessons found in
the wild. A companion document to this, [lessondown-example-blocks.Rmd] should
contain RMarkdown blocks that ideally should be able to translate to these
markdown blocks.

The way the blocks are currently set up (the `{: .challenge}` tag always appears
after block quotes), makes it relatively straightforward to grab them using
XPATH syntax:

```{r, eval = FALSE}
library("magrittr")
library("xml2")
library("tinkr") # https://github.com/ropenscilabs/tinkr
library("tibble")
library("purrr")
library("stringr")
library("git2r")
library("fs")


lesson <- to_xml("/path/to/_episodes/lesson.md")
lesson$body %>%
  xml_find_all(".//d1:text[text()='{: .challenge}']/ancestor-or-self::d1:block_quote")
```

At the moment, I'm not quite sure how to extract this information in a useable
manner since XML is represented as pointers. using `xml_text()` will collapse
all child nodes a single text field, delimted by nothing, so it comes out weird:

```{r, eval = FALSE}
lesson$body %>%
  xml_find_first(".//d1:text[text()='{: .challenge}']/ancestor-or-self::d1:block_quote") %>%
  xml_text()
# [1] "Fill in the BlanksFill in the blanks so that the program below produces the output shown.values = ____\nvalues.____(1)\nvalues.____(3)\nvalues.____(5)\nprint('first time:', values)\nvalues = values[____]\nprint('second time:', values)\n{: .language-python}first time: [1, 3, 5]\nsecond time: [3, 5]\n{: .output}Solutionvalues = []\nvalues.append(1)\nvalues.append(3)\nvalues.append(5)\nprint('first time:', values)\nvalues = values[1:]\nprint('second time:', values)\n{: .language-python}{: .solution}{: .challenge}"
```

However, if we unnest things with `xml_contents()`, then we get printing of child nodes:

```{r, eval = FALSE}
lesson$body %>%
  xml_find_first(".//d1:text[text()='{: .challenge}']/ancestor-or-self::d1:block_quote") %>%
  xml_contents() -> blocks
xml_text(blocks)
# [1] "Fill in the Blanks"
# [2] "Fill in the blanks so that the program below produces the output shown."
# [3] "values = ____\nvalues.____(1)\nvalues.____(3)\nvalues.____(5)\nprint('first time:', values)\nvalues = values[____]\nprint('second time:', values)\n"
# [4] "{: .language-python}"
# [5] "first time: [1, 3, 5]\nsecond time: [3, 5]\n"
# [6] "{: .output}"
# [7] "Solutionvalues = []\nvalues.append(1)\nvalues.append(3)\nvalues.append(5)\nprint('first time:', values)\nvalues = values[1:]\nprint('second time:', values)\n{: .language-python}{: .solution}{: .challenge}"
```

Here, we are missing the names of the nodes, but we can get those with `xml_name()`:
```{r, eval = FALSE}
xml_name(blocks)
# [1] "heading"     "paragraph"   "code_block"  "paragraph"   "code_block"  "paragraph"
# [7] "block_quote"
tibble(node = xml_name(blocks), text = xml_text(blocks))
```

Update: there was a problem in my thinking because I realized that this search
was also including block quote children of the top-level block quotes (that is,
it was including the solutions twice). What we want are simply all of the
challenge block quotes.

I've written a functon that will search exactly for these block quotes and return
a nested list of the contents of said block quotes!

```{r, eval = TRUE}
peek <- function(x, n = 1, ...) str(x, max.level = n, ...)
carpentries_examples <- function(body, list = TRUE) {
  # Setting up the XPATH search string
  challenge <- "d1:text[text()='{: .challenge}']"       # Find the end of the challenge block
  axis      <- "ancestor-or-self"                       # Then look behind at all of the ancestors
  ancestor  <- "d1:block_quote"                         # That are blockquotes
  predicate <- "d1:heading/d1:text[text()!='Solution']" # But exclude the Solution blocks (because they are included anyways)

  challenge_string <- glue::glue(".//{challenge}/{axis}::{ancestor}[{predicate}]")
  examples <- xml2::xml_find_all(body, challenge_string)
  if (list) {
    examples <- xml2::as_list(examples)
  }
  examples
}
```

Now, if we peek at these examples, we can find out a couple of things:

1. How are they structured from a broad level (paragraph, code block, paragraph,
   list, solution (block_quote))
2. What are the text types that we need to look for at the elemental level?

```{r, eval = FALSE}
# Broad level
examples <- carpentries_examples(lesson$body)
lesson_names <- lapply(examples, names)
lesson_names
table(unlist(lesson_names))

# elemental level
element_names <- names(unlist(examples))
str_extract(element_names, "[_a-z]+?$") %>% table()
```

The question is: what happens when we do this for an entire lesson:

```{r}
lesson <- "https://github.com/swcarpentry/python-novice-gapminder.git"

if (!fs::dir_exists("swc--png")) {
  x <- git2r::clone(lesson, "swc--png")
}
episodes <- fs::dir_ls("swc--png/_episodes")

bodies <- map(episodes, ~to_xml(.x)$body %>% carpentries_examples(list = FALSE))
bodies <- bodies[lengths(bodies) > 0]
elist  <- map(bodies, as_list)

# How deep are the tags?
map_int(elist, vec_depth)
#          swc--png/_episodes/01-run-quit.md         swc--png/_episodes/02-variables.md
#                                          6                                          8
#  swc--png/_episodes/03-types-conversion.md          swc--png/_episodes/04-built-in.md
#                                          9                                         10
#         swc--png/_episodes/06-libraries.md   swc--png/_episodes/07-reading-tabular.md
#                                          8                                          8
#       swc--png/_episodes/08-data-frames.md          swc--png/_episodes/09-plotting.md
#                                          8                                          7
#             swc--png/_episodes/11-lists.md         swc--png/_episodes/12-for-loops.md
#                                          9                                          8
#      swc--png/_episodes/13-conditionals.md swc--png/_episodes/14-looping-data-sets.md
#                                          6                                          7
# swc--png/_episodes/16-writing-functions.md             swc--png/_episodes/17-scope.md
#                                          9                                          7
#             swc--png/_episodes/18-style.md
#                                          7

# What top level tags are there in the challenges?
map_depth(elist, .depth = 2, .f = names) %>% unlist() %>% table()
# .
# block_quote  code_block     heading  html_block        list   paragraph
#          76          78          76           1          28         179

# What top level tags are in the solutions?
map_depth(elist, .depth = 2, .f = ~names(.x[["block_quote"]])) %>% unlist() %>% table()
# .
# code_block    heading       list  paragraph      table
#         89         72         19        149          1

# What are the common patterns among the solutions (excluding duplicated)?

nodup <- function(x) x[c(x[-length(x)] != x[seq_along(x)[-1]], TRUE)]

patterns <- map(elist, ~map_chr(.x, .f = ~paste(nodup(names(.x[["block_quote"]])), collapse = "."))) %>%
  unlist() %>%
  unique() %>%
  str_remove(r"{^heading\.}")

patterns[order(nchar(patterns))]
#  [1] ""
#  [2] "list"
#  [3] "table"
#  [4] "paragraph"
#  [5] "paragraph.list"
#  [6] "code_block.paragraph"
#  [7] "paragraph.list.paragraph"
#  [8] "paragraph.code_block.paragraph"
#  [9] "code_block.paragraph.code_block"
# [10] "paragraph.code_block.paragraph.list"
# [11] "code_block.paragraph.code_block.paragraph"
# [12] "paragraph.code_block.paragraph.code_block.paragraph"
# [13] "code_block.paragraph.code_block.paragraph.code_block.paragraph"
# [14] "paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph"
# [15] "list.code_block.paragraph.list.code_block.paragraph.list.code_block.paragraph"
# [16] "paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph"
# [17] "paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph"
# [18] "list.code_block.paragraph.list.code_block.paragraph.list.paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph"
# [19] "list.code_block.paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph"
# [20] "code_block.paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph.code_block.paragraph"

# This makes me wonder what this would look like in graph form.

make_edges <- function(nms, src = "CHALLENGE", sink = "LESSON") {
  nms <- nms[nms != "heading"]
  data.frame(
    from = c(src, nms),
    to   = c(nms, sink)
  )
}

solution_graph <- map_dfr(elist,
  .f = ~map_dfr(.x, ~make_edges(names(.x[["block_quote"]]), src = "SOLUTION", sink = "CHALLENGE"), .id = "block"),
  .id = "episode"
) %>%
  dplyr::select(from, to, everything())

challenge_graph <- map_dfr(elist,
  .f = ~map_dfr(.x, ~make_edges(names(.x)), .id = "block"),
  .id = "episode"
) %>%
  dplyr::select(from, to, everything()) %>%
  dplyr::mutate(to = dplyr::if_else(to == "block_quote", "SOLUTION", to)) %>%
  dplyr::mutate(from = dplyr::if_else(from == "block_quote", "SOLUTION", from))

solution_graph %>% graph_from_data_frame() %>% simplify() %>%
  plot(layout = layout_in_circle(.), edge.curved = -0.25)
challenge_graph %>% graph_from_data_frame() %>% simplify() %>%
  plot(layout = layout_in_circle(.), edge.curved = -0.25)

dplyr::bind_rows(challenge_graph, solution_graph) %>%
  graph_from_data_frame() %>%
  set_edge_attr("weight", value = 1) %>%
  delete_edge_attr("episode") %>%
  delete_edge_attr("block") %>%
  simplify(remove.loops = FALSE, edge.attr.comb = sum) %>%
  plot(layout = layout_in_circle(.), edge.curved = -0.25, edge.width = sqrt(E(.)$weight))

# Now that we know the overal structure of a given lesson, what are the text-like attributes?
elist %>% unlist() %>% names() %>% str_extract("[_a-z]+?$") %>% table()
# .
#        code  code_block  html_block html_inline        text
#         304         170           1          18        1307

# Note that the only thing about this that it removes all of the softbreaks that exist.
```



