---
title: I don't know much about namespaces
author: Zhian N. Kamvar
---


I don't know much about namespaces in {xml2}, but I'm trying to figure it out. 
From what I can tell, creating and appending xml documents is fairly
straightforward (once you are comfortable with the structure). Here, I can
create a document that contains a node 'a' and nest a node 'b' underneath:

```{r first}
library("xml2")
x <- read_xml("<a></a>")
xml_add_child(x, read_xml("<b></b>"))
xml_structure(x)
xml_find_all(x, ".//b")
```

A more concrete example is the CD catalogue example. Let's say I wanted to make
sure that Santana's 'Supernatural' made the cut because life without 'Smooth'
by Santana feat. Rob Thomas from Matchbox 20 is no life at all.

```{r cd}
cd <- read_xml(xml2_example("cd_catalog.xml"))
xml_find_all(cd, ".//TITLE[text()='Supernatural']/parent::*")
smooth <- read_xml("<CD>
  <TITLE>Supernatural</TITLE>
  <ARTIST>Santana</ARTIST>
  <COMPANY>Fantasy Studios</COMPANY>
  <PRICE>priceless</PRICE>
  <YEAR>1999</YEAR>
</CD>")

xml_add_child(cd, smooth, .where = 0)
xml_find_all(cd, ".//TITLE[text()='Supernatural']/parent::*")
```

That's great! But what happens if smooth happens to have a namespace


```{r ns}
cd <- read_xml(xml2_example("cd_catalog.xml"))
smooth <- read_xml("<CD xmlns='http://featur.ing/rob-thomas-of-matchbox-20'>
  <TITLE>Supernatural</TITLE>
  <ARTIST>Santana</ARTIST>
  <COMPANY>Fantasy Studios</COMPANY>
  <PRICE>priceless</PRICE>
  <YEAR>1999</YEAR>
</CD>")
xml_add_child(cd, smooth, .where = 0)
xml_find_all(cd, ".//TITLE[text()='Supernatural']/parent::*")
xml_find_all(cd, ".//TITLE[text()='Greatest Hits']/parent::*")
```

All of a sudden, I can't find SMOOTH :(
If I use `xml_ns()`, I can find out what the name of the namespace is. 

```{r nsxpath}
(sns <- xml_ns(smooth))
xml_find_all(cd, ".//d1:TITLE[text()='Supernatural']/parent::*")
xml_find_all(cd, ".//d1:TITLE[text()='Greatest Hits']/parent::*")
```

So, I can get the name of the namespace, but I can't use it across the document
because the whole document doesn't have a namespace. I would use 
`xml_set_namespace()`, but the function requires that a node has some ancestor
with the same namespace.

```{r}
xml_attr(smooth, "xmlns")
xml_attr(cd, "xmlns")
```

What if I set the namespace using the attribute?

```{r}
xml_set_attr(cd, "xmlns", xml_attr(smooth, "xmlns"))
xml_find_all(cd, ".//d1:TITLE[text()='Supernatural']/parent::*")
xml_find_all(cd, ".//d1:TITLE[text()='Greatest Hits']/parent::*")
xml_ns(cd)
```

It worked! Now let's try the situation where I insert a node with no namespace
into a document that has a namespace

```{r}
cd <- read_xml(xml2_example("cd_catalog.xml"))
xml_set_attr(cd, "xmlns", xml_attr(smooth, "xmlns"))
smooth <- read_xml("<CD>
  <TITLE>Supernatural</TITLE>
  <ARTIST>Santana</ARTIST>
  <COMPANY>Fantasy Studios</COMPANY>
  <PRICE>priceless</PRICE>
  <YEAR>1999</YEAR>
</CD>")
xml_add_child(cd, smooth, .where = 0)
xml_find_all(cd, ".//d1:TITLE[text()='Supernatural']/parent::*")
xml_find_all(cd, ".//d1:TITLE[text()='Greatest Hits']/parent::*")
```

AHA! This is one of the problems I've been having, if you have a document that
starts out with a namespace, and you insert nodes without that namespace, then
you will have a mismatch and you can't select those nodes. What if we used
`xml_set_namespace()` on this?

```{r}
xml_set_namespace(xml_child(cd), "d1", xml_ns(cd))
xml_find_all(cd, ".//d1:TITLE[text()='Supernatural']/parent::*")
xml_find_all(cd, ".//d1:TITLE[text()='Greatest Hits']/parent::*")
```

Well, that's frustrating. What's even more frustrating is that xml_ns reports
that the namespace has already been set. Maybe we can add it via attributes

```{r}
xml_set_attr(xml_child(cd), "xmlns", xml_ns(cd))
xml_find_all(cd, ".//d1:TITLE[text()='Supernatural']/parent::*")
xml_find_all(cd, ".//d1:TITLE[text()='Greatest Hits']/parent::*")
```

YES! But now, we have that attribute hanging there and it's weird. 

# Why is this relevant?

The namespace of nodes is relevant for me at the very least because it affects
how the document is processed by XSLT: https://www.w3schools.com/XML/xml_namespaces.asp

It's of concern because the {tinkr} package processes markdown documents via
the {commonmark} package. For example:

```{r common}
library(commonmark)
cat(lst <- markdown_xml("- eggs\n- butter\n- milk"))
x <- read_xml(lst)
```

The document has a namespace that points to "http://commonmark.org/xml/1.0" and
that's used in the [stylesheet for {tinkr}](https://github.com/ropenscilabs/tinkr/blob/4980e816d876a86bb1ea07fe2b564701bc4bce22/inst/extdata/xml2md_gfm.xsl#L4-L5).

If an item is in the commonmark namespace, it will be processed by the
stylesheet. Tinkr does this via the {xslt} package, which provides an interface to the tempaltes and outputs text:

```{r xslt}
library(xslt)
stylesheet <- read_xml(system.file("extdata", "xml2md_gfm.xsl", package = "tinkr"))
cat(xslt::xml_xslt(x, stylesheet))
xlist <- xml_find_first(x, ".//d1:list")
xlist
```

```{r}
brandy <- read_xml("<item><paragraph><text>brandy</text></paragraph></item>")
xml_add_child(xlist, brandy)
xlist
xml_find_all(x, ".//d1:text[text()='brandy']")
xml_find_all(x, ".//d1:text[text()='butter']")
# Adding this child has no effect
cat(xslt::xml_xslt(x, stylesheet))
```

So, what can we do? We can try adding a namspace to the node


```{r}
xml_set_attr(xml_children(xlist), "xmlns", xml_ns(x))
xlist
cat(xslt::xml_xslt(x, stylesheet))
```

It works! But that is a bit of a hassle because now everything shows up with
that xmlns attribute:

```{r}
xml_attrs(xml_children(xlist))
```

What if we defined the namespace from the get-go?


```{r}
x <- read_xml(lst)
ns <- glue::glue("xmlns='{xml_ns(x)}'")
brandy <- read_xml(glue::glue("<item {ns}><paragraph><text>brandy</text></paragraph></item>"))
xlist <- xml_find_first(x, ".//d1:list")
xml_add_child(xlist, brandy)
xlist
xml_find_all(x, ".//d1:text[text()='brandy']")
xml_find_all(x, ".//d1:text[text()='butter']")
# Adding this child has no effect
cat(xslt::xml_xslt(x, stylesheet))
xml_ns(x)
```


IT WORKS!!!!

Let's see if it will work with tinkr. I'm going to add a code block at the top
of a README:

```{r}
x <- tinkr::to_xml(system.file("extdata", "example2.Rmd", package = "tinkr"))

# It's a bit weird, but there must be a newline character after the code
make_code_block <- function(code = "print('hello world?')\n", name = 'cody', body) {
  block <- glue::glue("
    <document xmlns='{xml2::xml_ns(body)[[1]]}'>
      <code_block language='r' name='{name}' echo='TRUE' eval='TRUE'>{code}</code_block>
    </document>"
  )
  xml2::xml_child(xml2::read_xml(block))
}

# insert a new code block after the setup chunk
xml2::xml_add_child(x$body, make_code_block(body = x$body), .where = 1)

# find that code block
xml2::xml_find_all(x$body, ".//d1:code_block[@name='setup']")
xml2::xml_find_all(x$body, ".//d1:code_block[@name='cody']")

# Round trip does not work
tou <- tempfile(fileext = ".Rmd")
tinkr::to_md(x, tou)
xml2::xml_find_all(tinkr::to_xml(tou)$body, ".//d1:code_block[@name='setup']")
xml2::xml_find_all(tinkr::to_xml(tou)$body, ".//d1:code_block[@name='cody']")
```

As of this writing (2020-09-23) this failed spectacularly. Let's see if we can
break away from the {tinkr} package and focus on {commonmark} and {xslt}:


```{r}
stylesheet <- read_xml(system.file("extdata", "xml2md_gfm.xsl", package = "tinkr"))
message(md <- "# hello\n\n`code` text\n\n```{r}\ncat('a code block')\n```")
message(xmd <- commonmark::markdown_xml(md))
x <- xml2::read_xml(xmd)
# round trip from markdown -> xml -> markdown
message(xslt::xml_xslt(x, stylesheet))
```

Now, let's see what happens if we try to insert a new code block into the mix

```{r}
print(cb <- make_code_block(body = x))
xml2::xml_add_child(x, cb, .where = 1)
message(xslt::xml_xslt(x, stylesheet))
```

It's  a bit weird, but the text of a code block MUST have a newline at the end
or else, it won't appear:


```{r}
print(cb <- make_code_block("# This comment shows up\n# This comment is hidden", body = x))
xml2::xml_add_child(x, cb, .where = 1)
message(xslt::xml_xslt(x, stylesheet))
```

I've addressed this in https://github.com/ropenscilabs/tinkr/pull/24










